Final draft

### 2021-07-15 final draft.Rmd
space
SV, adding comment on density forecasts
piccole correzioni

---
header-includes:
- \usepackage{txfonts}
- \usepackage{amsmath}
- \usepackage[bindingoffset=1.5cm, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
- \usepackage{graphicx}
- \usepackage{MnSymbol} %allows to insert the QED at the end of the proofs
- \usepackage{hyperref} %set the hyperlink to cite sections, equations...
- \urlstyle{same}
- \usepackage{comment}
- \usepackage{enumitem}
- \usepackage{ntheorem}
- \usepackage{lipsum}
- \usepackage{amssymb}
- \usepackage{xcolor}
- \theoremstyle{break}
- \newtheorem{thm}{Theorem}% theorem counter resets every \subsection
- \renewcommand{\thethm}{\arabic{thm}}
- \newtheorem{proposition}{Proposition}
- \newtheorem{assumption}{Assumption}
- \newtheorem{lemma}{Lemma}
- \newtheorem{definition}{Definition}
- \theoremstyle{nonumberplain}
- \newtheorem{proof*}{Proof.}
- \newtheorem{definition}{Definition}
- \renewcommand{\P}{\mathbb{P}}
- \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
- \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
- \setcounter{MaxMatrixCols}{10}
- \setlength{\oddsidemargin}{.0in}
- \setlength{\textwidth}{6 in}
- \setlength{\topmargin}{-.1in}
- \setlength{\textheight}{8.1in}
- \renewcommand{\baselinestretch}{1.5}
- \usepackage{amssymb}
- \newcommand{\E}{\mathbb E}
- \newcommand{\Q}{\mathbb Q}
- \newcommand{\M}{\mathbb M}
- \newcommand{\R}{\mathbb R}
- \newcommand{\B}{\mathcal B}
- \newcommand{\X}{\mathcal X}
- \newcommand{\Pd}{\mathbb P}
- \newcommand{\N}{\mathcal N}
- \def\code#1{\texttt{#1}}
- \newtheorem{algorithm}{Algorithm}[section]
- \usepackage{float}
- \floatplacement{figure}{}

@@

Let $P_1, P_2, \dots, P_T$ be a finite measure of probability kernels from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ Fix some $\mathbb{P}_0(dx)$ on $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ \begin{definition}
A discrete-time Markov process is a sequence $X_{0:T}$ of random variables whose joint distribution can be written as
\begin{equation}
    \P_T(X_{0:T}\in dx_{0:T})=\P_0(dx_0)\prod\limits_{s=1}^T P_s(x_{s-1},x_s).
\end{equation}
The set $\mathcal{X}$ is the state-space, $\P_0$ is the initial distribution and the probability kernel $P_t$ is the transition kernel.
\end{definition}
It is possible to show that for any $t\in T,$
$$
\P_T(X_t\in dx_t\mid X_{0:t-1}\in dx_{0:t-1})=
\P_T(X_t\in dx_t\mid X_{t-1}=x_{t-1})=\P_t(x_{t-1},dx_t).
$$
The first equality implies conditional independence. The second equality implies that it is possible to identify conditional distributions with transition kernels. 
Assume $\P_t, t\leq T,$ is a sequence of probability measures. Then, the following proposition holds.
\begin{proposition}
For all $t\leq T,$ 
$$
\P_T(dx_{0:t})=\P_0(dx_0)\prod\limits_{s=1}^t P_s(x_{s-1},dx_s)=\P_t(dx_{0:t}).
$$
\end{proposition}
This means that the marginal distribution of $X_{0:t}$ with respect to $\P_T$ is $\P_t.$

\section{State-Space Models}
We now define \textit{State-Space models} as Markov processes that are partially observed. Let $X_t\in\X$ and $Y_t\in\Y$ and let $\{(X_t,Y_t)\}$ be a stochastic process. To help intuition, suppose that $\{Y_t\}$ is observable, i.e., coincides with a sequence of observations, while $\{X_t\}$ is latent. Consider the initial distribution $\P_0(dx_0)$ and the probability kernels
$P_t(x_{t-1},dx_t)$ and $F_t(x_t,d_t)$, where $t=1,2,\dots$.
\begin{definition}\label{state-space-def}
The process $\{(X_t,Y_t)\}$ is a State-Space model if the joint distribution of $(X_{0:T},Y_{0:T})$ is 
\begin{align*}
  \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T}) &= \P_0(dx_0)\prod\limits_{t=1}^T P_t(x_{t-1},dx_t)\prod\limits_{t=1}^T F_t(x_t,dy_t)\\
  &= [\P_0(dx_0)F_0(x_0,y_0)]\prod\limits_{t=1}^T[P_t(x_{t-1},dx_t)F_t(x_t,dy_t)] \\
  &= \P_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t)].
\end{align*}
\end{definition}
From the second inequality, it is apparent that $\{(X_t,T_t\}$ is a Markov process with initial distribution $[\P_0(dx_0)F_0(x_0,y_0)]$ and transition kernel $P_t(x_{t-1},dx_t)F_t(x_t,dy_t).$ The third equality, instead, shows that, $\{X_t\}$ has a marginal distribution $\P_T(dx_{0:T})=\P_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),$ meaning that $\{X_t\}$ is marginally distributed as a Markov process with initial distribution $\P_T(dx_{0:T})$ and transition kernel $P_t(x_{t-1},dx_t).$

Assume now that there exist a dominating measures $\nu(dy)$ such that 
$F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)$ for all $t.$
Then by the last equality of definition \ref{state-space-ref}, the joint distribution of $(X_{0:T},Y_{0:T})$ is 
\begin{equation}\label{state-space-densities}
   \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\P_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\vu(dy_t).
\end{equation}
Integrating away $X_{0:T},$ one gets the marginal density
\begin{align*}\label{state-space-marginal-y}
  \P_T(dy_{0:T}) &= \E_{\P_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\vu(dy_s) \\
                &= p_t(y_{0:t})\prod\limits_{s=0}^t\vu(dy_s),
\end{align*}
where $p_t(y_{0:t})$ is the likelihood function.
Dividing the left-hand side of equation \ref{state-space-densities} by the left-hand side of equation \ref{state-space-marginal-y} and the right-hand side of equation \ref{state-space-densities} by the right-hand side of equation \ref{state-space-marginal-y}, one obtains the filtering distribution
\begin{equation}
\P_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\P_t(dx_{0:t}).
\end{equation}

@@

Suppose that $\X=\R^{d_x}$ and $\Y=\R^{d_y}.$ Let $U_t\thicksim\N(0,\Sigma_t)$ and $V_t\thicksim\N(0,R_t)$ be two independent white-noise errors and assume the initial distribution 
$X_0\thicksim\N(0,\Sigma_0).$ We have a linear Gaussian model if 
\begin{align}
X_t &= A_tX_{t-1}+U_t \\
Y_t &= B_tX_t + V_t.
\end{align}
Note that  $X_t\mid X_{t-1}$ and $Y_t\mid X_t$ are both normally distributed. 
\begin{lemma}\label{kalman-filter}
  Let $X\thicksim\N(m_0,Q_0)$ and $Y\mid X\thicksim\N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
  Then, it holds that
  \begin{align}
    X\mid Y=y&\thicksim\N(m_1,Q_1),\\
    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
  \end{align}
\end{lemma}
\begin{proof}

\end{proof}

Assume that $\P_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\N(m_{t-1},Q_{t-1}).$
Then, using the transition equation of the latent variable, 
by the linearity of the Gaussian distribution, we have that the predictive distribution is $\P_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).$
Letting $E_T=A_tQ_{t-1}A_t'+\Sigma_t,$ applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
\begin{align*}
  \P_{t}(dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
\end{align*}

>>>>

---
header-includes:
- \usepackage{txfonts}
- \usepackage{amsmath}
- \usepackage[bindingoffset=1.5cm, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
- \usepackage{graphicx}
- \usepackage{MnSymbol} %allows to insert the QED at the end of the proofs
- \usepackage{hyperref} %set the hyperlink to cite sections, equations...
- \urlstyle{same}
- \usepackage{comment}
- \usepackage{enumitem}
- \usepackage{ntheorem}
- \usepackage{lipsum}
- \usepackage{amssymb}
- \usepackage{xcolor}
- \theoremstyle{break}
- \newtheorem{thm}{Theorem}% theorem counter resets every \subsection
- \renewcommand{\thethm}{\arabic{thm}}
- \newtheorem{proposition}{Proposition}
- \newtheorem{assumption}{Assumption}
- \newtheorem{lemma}{Lemma}
- \newtheorem{definition}{Definition}
- \theoremstyle{nonumberplain}
- \newtheorem{proof*}{Proof.}
- \newtheorem{definition}{Definition}
- \renewcommand{\P}{\mathbb{P}}
- \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
- \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
- \setcounter{MaxMatrixCols}{10}
- \setlength{\oddsidemargin}{.0in}
- \setlength{\textwidth}{6 in}
- \setlength{\topmargin}{-.1in}
- \setlength{\textheight}{8.1in}
- \renewcommand{\baselinestretch}{1.5}
- \usepackage{amssymb}
- \newcommand{\E}{\mathbb E}
- \newcommand{\Q}{\mathbb Q}
- \newcommand{\M}{\mathbb M}
- \newcommand{\R}{\mathbb R}
- \newcommand{\B}{\mathcal B}
- \newcommand{\X}{\mathcal X}
- \newcommand{\Y}{\mathcal Y}
- \newcommand{\Pd}{\mathbb P}
- \newcommand{\N}{\mathcal N}
- \def\code#1{\texttt{#1}}
- \newtheorem{algorithm}{Algorithm}[section]
- \usepackage{float}
- \floatplacement{figure}{}

@@

Let $P_1, P_2, \dots, P_T$ be a finite sequence of probability kernels from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ Fix some $\mathbb{P}_0(dx)$ on $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ \begin{definition}
A discrete-time Markov process is a sequence $X_{0:T}$ of random variables whose joint distribution can be written as
\begin{equation}
    \P_T(X_{0:T}\in dx_{0:T})=\P_0(dx_0)\prod\limits_{s=1}^T P_s(x_{s-1},dx_s).
\end{equation}
The set $\mathcal{X}$ is the state-space, $\P_0$ is the initial distribution and the probability kernel $P_t$ is the transition kernel.
\end{definition}
It is possible to show that for any $t\in T,$
$$
\P_T(X_t\in dx_t\mid X_{0:t-1}\in dx_{0:t-1})=
\P_T(X_t\in dx_t\mid X_{t-1}=x_{t-1})=\P_t(x_{t-1},dx_t).
$$
The first equality implies conditional independence. The second equality implies that it is possible to identify conditional distributions with transition kernels. 
Assume $\P_t,$ with $t\leq T,$ is a sequence of probability measures. Then, the following proposition holds.
\begin{proposition}
For all $t\leq T,$ 
$$
\P_T(dx_{0:t})=\P_0(dx_0)\prod\limits_{s=1}^t P_s(x_{s-1},dx_s)=\P_t(dx_{0:t}).
$$
\end{proposition}
This means that the marginal distribution of $X_{0:t}$ with respect to $\P_T$ is $\P_t.$

\section{State-Space Models}
We now define \textit{State-Space models} as Markov processes that are partially observed. Let $X_t\in\X$ and $Y_t\in\Y$ and let $\{(X_t,Y_t)\}$ be a stochastic process. To help intuition, suppose that $\{Y_t\}$ is observable, i.e., coincides with a sequence of observations, while $\{X_t\}$ is latent. Consider the initial distribution $\P_0(dx_0)$ and the probability kernels
$P_t(x_{t-1},dx_t)$ and $F_t(x_t,d_t)$, where $t=1,2,\dots$.
\begin{definition}\label{state-space-def}
The process $\{(X_t,Y_t)\}$ is a State-Space model if the joint distribution of $(X_{0:T},Y_{0:T})$ is 
\begin{align*}
  \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T}) &= \P_0(dx_0)\prod\limits_{t=1}^T P_t(x_{t-1},dx_t)\prod\limits_{t=1}^T F_t(x_t,dy_t)\\
  &= [\P_0(dx_0)F_0(x_0,y_0)]\prod\limits_{t=1}^T[P_t(x_{t-1},dx_t)F_t(x_t,dy_t)] \\
  &= \P_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t).
\end{align*}
\end{definition}
From the second inequality, it is apparent that $\{(X_t,Y_t)\}$ is a Markov process with initial distribution $[\P_0(dx_0)F_0(x_0,y_0)]$ and transition kernel $P_t(x_{t-1},dx_t)F_t(x_t,dy_t).$ The third equality, instead, shows that, $\{X_t\}$ has a marginal distribution $$\P_T(dx_{0:T})=\P_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),$$ meaning that $\{X_t\}$ is marginally distributed as a Markov process with initial distribution $\P_T(dx_{0:T})$ and transition kernel $P_t(x_{t-1},dx_t).$

Assume now that there exists a dominating measures $\nu(dy)$ such that 
$F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)$ for all $t.$
Then by the last equality of definition \ref{state-space-def}, the joint distribution of $(X_{0:T},Y_{0:T})$ is 
\begin{equation}\label{state-space-densities}
   \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\P_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\vu(dy_t).
\end{equation}
Integrating away $X_{0:T},$ one gets the marginal density
\begin{equation}\label{state-space-marginal-y}
\begin{align*}
  \P_T(dy_{0:T}) &= \E_{\P_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
\end{align*}                
\end{equation}
where $p_t(y_{0:t})$ is the likelihood function.
Dividing the left-hand side of equation \ref{state-space-densities} by the left-hand side of equation \ref{state-space-marginal-y} and the right-hand side of equation \ref{state-space-densities} by the right-hand side of equation \ref{state-space-marginal-y}, one obtains the filtering distribution
\begin{equation}
\P_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\P_t(dx_{0:t}).
\end{equation}

@@

Suppose that $\X=\R^{d_x}$ and $\Y=\R^{d_y}.$ Let $U_t\thicksim\N(0,\Sigma_t)$ and $V_t\thicksim\N(0,R_t)$ be two independent white-noise errors and assume the initial distribution 
$X_0\thicksim\N(0,\Sigma_0).$ We have a linear Gaussian model if 
\begin{align}
X_t &= A_tX_{t-1}+U_t \\
Y_t &= B_tX_t + V_t.
\end{align}
Note that  $X_t\mid X_{t-1}$ and $Y_t\mid X_t$ are both normally distributed. 
\begin{lemma}\label{kalman-filter}
  Let $X\thicksim\N(m_0,Q_0)$ and $Y\mid X\thicksim\N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
  Then, it holds that
  \begin{align}
    X\mid Y=y&\thicksim\N(m_1,Q_1),\\
    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
  \end{align}
\end{lemma}

Assume that $\P_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\N(m_{t-1},Q_{t-1}).$
Then, using the transition equation of the latent variable, 
by the linearity of the Gaussian distribution, we have that the predictive distribution is 
\begin{equation}
\P_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
\end{equation}
Letting $E_T=A_tQ_{t-1}A_t'+\Sigma_t,$ and applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
\begin{align*}
  \P_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
\end{align*}
